6.829 Final Project: A comparison of bit rate selection algorithms

### 2013-05-17 paper/paper.tex
readme adjustments and fixed typo

Minstrel makes use the multi-rate retry chain (MRR), an array of bit rates
and number of attempts to make at each bit rate, that tells the card
which rates to try before reporting a failure.  The retry chain makes
failures extremely unlikely and allows Minstrel to choose fallback
rates when a packet does not succeed.  Minstrel sets the MRR to first
try the rate with the highest throughput, then the next-highest
throughput, then greatest probability of success, and finally attempts
to send a packet at the lowest, base rate.  In essence, Minstrel tells
the card to send at the highest throughput rates but to fall back to
reliable rates upon failure.

@@

We implemented SampleRate as outlined in John Bicket's Master's
thesis~\cite{samplerate}.  This implementation is slightly different
from how it was implemented in the MadWifi kernel driver.  The primary
difference is that the kernel implementatiion uses a EWMA, while the
thesis implementation computes average transmission times based on a
10-second sliding window.  It would be worth looking at the
performance of EWMA-based SampleRate in the future.  Our
implementation, \texttt{samplerate.py} contains \texttt{ApplyRate()},
\texttt{ProcessFeedback()} and \texttt{RemoveStaleResults()}, as well
as a few helper functions and data structures for tracking rate
statistics.

@@

Minstrel and SampleRate perform similarly, usually within 3 Mbps of
each other.  In especially noisy situations, such as when the client
is moving or when the client is around the corner from the access
point, Minstrel consistently outperforms SampleRate.
Figure~\ref{figure:3} shows the results of a single run of both
Minstrel and SampleRate over all the different types of traces we
collected.  This is consisten with the literature---the authors of
Minstrel noted that they focused on making Minstrel robust to poor
conditions~\cite{minstrel}, which means losing potential througput
gains in stable situations.  Conversely, John Bicket noted that
SampleRate performs better in stable situations than noisy or highly
varying conditions~\cite{samplerate}.  We were slightly surprised to
see how significantly SampleRate outperformed Minstrel in certain
situations, though, with throughput being up to $67\%$ higher.

@@

In Figure~\ref{figure:2} we show the performance of constant bit rates
on the \texttt{clear\_1.dat} trace, which is a 35 second trace in
clear view of the access point.  The optimal rate here was 18 Mbps,
which achieved an average throughput of about 15 Mbps.  The next
highest bit rate, 24 Mbps, shows a large plummet in performance, down
to a throughput of 1.17 Mbps.  This steep drop is charateristic, and
is also talked about in the SampleRate thesis~\cite{samplerate}.  It
demonstrates just how important bit rate selection is---the optimal
bit rate has typically much higher throughput than the next-best rate,
but if the algorithm strays too high the results can be devastating.

@@

The histograms of which rates the algorithms decided to send at
provided some useful insight into their behavior.  Table~\ref{table:2}
is a histogram of a 35 second corner trace.  In this situation
Minstrel outperforms SampleRate.  The constant rate throughputs
demonstrate that 11, 12 and 18 Mbps were the optimal rates to send at,
and that no packets were sucessful at any higher rates.  SampleRate
does not use these rates except for probe packets, which happens once
every ten seconds.  Minstrel, since it does not terminate transmission
after four successive failures, spends a lot of time sending probe
packets at these high rates, which are never successful.  The only
reason low rates have fewer probe packets is because Minstrel
explicitly places them second in the retry chain, and because the main
rate usually succeeds.  Despite the fact that SampleRate spends less
time probing high rates, it is less successful.  This is because
Minstrel send the most packets at the three most successful bit rates.
SampleRate, however, spends too much time sampling low bit rates.
Additionally, it spends a lot of time sending at 5.5 Mbps.  We suspect
that this might be because the equation SampleRate uses to estimate
average transmission time needs to be tweaked.
%   TODO: why does SampleRate spend time at 5.5 Mbps???

Our examination of the histograms showed us that Minstrel spends a lot
of time sampling at high rates.  In the specific case we examined, they
were all probe packets.  But sometimes Minstrel would choose 54Mbps as
the rate with the best throughput, which made little sense.  A closer
investigation revealed that Minstrel would sometimes get lucky with a
successful probe at 54Mbps.  The probability of success would be 1, so
then 54Mbps would be chosen as the rate with the best throughput.  This
is unfortunate, but the real harm comes in the next window: packets
would now be send at 54Mbps, and nearly all would fail.  However,
since these hundreds of failed packets weight equally with the single
successful probe packet, Minstrel still considers 54 Mbps the best
rate.  Thus the probability of success might drop from 1 to about 0.70,
which is still high enough for the throughput equation to rank 54 Mbps
highly.  Fundamentally, the problem is that the EWMA does not account
the second block having hundreds more packets than the first.  In the
next section we will talk about Minproved, an improved version of
Minstrel where we attempt to fix this problem.

>>>>

Minstrel makes use the multi-rate retry chain (MRR), an array of bit rates
and number of attempts to make at each bit rate, that tells the card
which rates to try before reporting a failure.  The retry chain makes
failures extremely unlikely and allows Minstrel to choose fall-back
rates when a packet does not succeed.  Minstrel sets the MRR to first
try the rate with the highest throughput, then the next-highest
throughput, then greatest probability of success, and finally attempts
to send a packet at the lowest, base rate.  In essence, Minstrel tells
the card to send at the highest throughput rates but to fall back to
reliable rates upon failure.

@@

We implemented SampleRate as outlined in John Bicket's Master's
thesis~\cite{samplerate}.  This implementation is slightly different
from how it was implemented in the MadWifi kernel driver.  The primary
difference is that the kernel implementation uses a EWMA, while the
thesis implementation computes average transmission times based on a
10-second sliding window.  It would be worth looking at the
performance of EWMA-based SampleRate in the future.  Our
implementation, \texttt{samplerate.py} contains \texttt{ApplyRate()},
\texttt{ProcessFeedback()} and \texttt{RemoveStaleResults()}, as well
as a few helper functions and data structures for tracking rate
statistics.

@@

Minstrel and SampleRate perform similarly, usually within 3 Mbps of
each other.  In especially noisy situations, such as when the client
is moving or when the client is around the corner from the access
point, Minstrel consistently outperforms SampleRate.
Figure~\ref{figure:3} shows the results of a single run of both
Minstrel and SampleRate over all the different types of traces we
collected.  This is consistent with the literature---the authors of
Minstrel noted that they focused on making Minstrel robust to poor
conditions~\cite{minstrel}, which means losing potential throughput
gains in stable situations.  Conversely, John Bicket noted that
SampleRate performs better in stable situations than noisy or highly
varying conditions~\cite{samplerate}.  We were slightly surprised to
see how significantly SampleRate outperformed Minstrel in certain
situations, though, with throughput being up to $67\%$ higher.

@@

In Figure~\ref{figure:2} we show the performance of constant bit rates
on the \texttt{clear\_1.dat} trace, which is a 35 second trace in
clear view of the access point.  The optimal rate here was 18 Mbps,
which achieved an average throughput of about 15 Mbps.  The next
highest bit rate, 24 Mbps, shows a large plummet in performance, down
to a throughput of 1.17 Mbps.  This steep drop is characteristic, and
is also talked about in the SampleRate thesis~\cite{samplerate}.  It
demonstrates just how important bit rate selection is---the optimal
bit rate has typically much higher throughput than the next-best rate,
but if the algorithm strays too high the results can be devastating.

@@

The histograms of which rates the algorithms decided to send at
provided some useful insight into their behavior.  Table~\ref{table:2}
is a histogram of a 35 second corner trace.  In this situation
Minstrel outperforms SampleRate.  The constant rate throughputs
demonstrate that 11, 12 and 18 Mbps were the optimal rates to send at,
and that no packets were successful at any higher rates.  SampleRate
does not use these rates except for probe packets, which happens once
every ten seconds.  Minstrel, since it does not terminate transmission
after four successive failures, spends a lot of time sending probe
packets at these high rates, which are never successful.  The only
reason low rates have fewer probe packets is because Minstrel
explicitly places them second in the retry chain, and because the main
rate usually succeeds.  Despite the fact that SampleRate spends less
time probing high rates, it is less successful.  This is because
Minstrel send the most packets at the three most successful bit rates.
SampleRate, however, spends too much time sampling low bit rates.
Additionally, it spends a lot of time sending at 5.5 Mbps.  We suspect
that this might be because the equation SampleRate uses to estimate
average transmission time needs to be tweaked.
%   TODO: why does SampleRate spend time at 5.5 Mbps???

Our examination of the histograms showed us that Minstrel spends a lot
of time sampling at high rates.  In the specific case we examined, they
were all probe packets.  But sometimes Minstrel would choose 54 Mbps as
the rate with the best throughput, which made little sense.  A closer
investigation revealed that Minstrel would sometimes get lucky with a
successful probe at 54 Mbps.  The probability of success would be 1, so
then 54 Mbps would be chosen as the rate with the best throughput.  This
is unfortunate, but the real harm comes in the next window: packets
would now be send at 54 Mbps, and nearly all would fail.  However,
since these hundreds of failed packets weight equally with the single
successful probe packet, Minstrel still considers 54 Mbps the best
rate.  Thus the probability of success might drop from 1 to about 0.70,
which is still high enough for the throughput equation to rank 54 Mbps
highly.  Fundamentally, the problem is that the EWMA does not account
the second block having hundreds more packets than the first.  In the
next section we will talk about Minproved, an improved version of
Minstrel where we attempt to fix this problem.
